### 2015-06-06 chapters/ipa_meets_unicode.tex
update IPA intro

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into some Latin-based characters
\citep{IPA2005}. Although IPA is reasonably easily adhered to with pen and
paper, it is not trivial to encode IPA characters electronically. Early work
addressing the need for a universal computing environment for writing systems
and their computational complexity is discussed in Simons 1989. For a long time,
linguists (like all other computer users) were limited to ASCII-encoded 7-bit
characters, which only includes Latin characters, numbers and some punctuation
and symbols. Restricted to these standard character sets that lacked IPA support
or other language-specific graphemes that they needed, linguists devised their
own solutions \citep{BirdSimons2003}. For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of to represent .
Tech-savvy linguists redefined selected characters from a character encoding by
mapping custom made fonts to those code points. However, one linguist's
electronic text would not render properly on another linguist's computer without
access to the same font. Further, if two character encodings defined two
character sets differently, then data could not be reliably and correctly
displayed. This is a common example of the non-interoperability of data and data
formats.

To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit
printable ASCII character sequences, e.g. <p\textbackslash> for [ɸ]. Two
problems with SAMPA are that (i) it is only a partial encoding of the IPA and
(ii) it encodes different languages in separate data tables, instead of a
universal alphabet, like IPA. SAMPA tables are derived from phonemes appearing
in several European languages that were developed as part of a European
Commission-funded project to address technical problems like electronic mail
exchange (what is now simply called email). SAMPA is essentially a hack to work
around displaying IPA characters, but it provided speech technology and other
fields a basis that has been widely adopted and used in code. So, SAMPA was a
collection of tables to be compared, instead of a large universal table
representing all languages. An extended version of SAMPA, called X-SAMPA, set
out to include every symbol in the IPA chart including all diacritics
\citep{WellsND}. X-SAMPA was considered more universally applicable because it
consisted of one table that encoded the set of characters that represented
phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used
for speech technology and as an encoding system in computational linguistics.
Eventually, ASCII-encoding of the IPA became depreciated through the advent of
the Unicode Standard. Note however that many popular software packages used for
linguistic analyses still require ASCII input,
e.g.~RuG/L04\footnote{\url{http://www.let.rug.nl/kleiweg/L04/}} and
SplitsTree4.\footnote{\url{http://www.splitstree.org/}}

Still, there are a few pitfalls to be aware of when using the Unicode Standard
to encode IPA. As we have said before, from a linguistic perspective it might
look like the Unicode Consortium is making incomprehensible decisions, but it is
important to realize that the consortium has tried and is continuing to try to
be as consistent as possible across a wide range of use cases, and it does place
linguistic traditions above other orthographic possibilities. In general, we
strongly suggest linguists not to complain about any decisions in the Unicode
Standard, but to try and understand the rationale of the Unicode Consortium
(which in our experience is almost always well-conceived) and devise ways to
work with any unexpected behavior. Many of the current problems derive from the
fact that the IPA is clearly historically based on the Latin script, but
different enough from most other Latin-based writing systems to warrant special
attention. This ambivalent status of the IPA glyphs (partly Latin, partly
special) is unfortunately also attested in the treatment of IPA in the Unicode
Standard. In retrospect, it might have been better to consider the IPA (and
other transcription systems) to be a special or new kind of script within the
Unicode Standard, and treat the obvious similarity to Latin glyphs as a
historical relic. All IPA glyphs would then have their own code points, instead
of the current situation in which some IPA glyphs have special code points,
while others are treated as being identical to the regular Latin characters.
Yet, the current situation, however unfortunate, is unlikely to change, so as
linguists we will have to learn to deal with the specific pitfalls of IPA within
the Unicode Standard. In this section, we will describe these pitfalls in some
detail.

>>>>

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into some Latin-based characters
\citep{IPA2005}. Although IPA is reasonably easily adhered to with pen and
paper, it is not trivial to encode IPA characters electronically. Early work
addressing the need for a universal computing environment for writing systems
and their computational complexity is discussed in \citet{Simons1989}. For a long
time, linguists (like all other computer users) were limited to ASCII-encoded
7-bit characters, which only includes Latin characters, numbers and some
punctuation and symbols. Restricted to these standard character sets that lacked
IPA support or other language-specific graphemes that they needed, linguists
devised their own solutions \citep{BirdSimons2003}. For example, some chose to
represent unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a character
encoding by mapping custom made fonts to those code points. However, one
linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Further, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a common example of the non-interoperability of
data and data formats.

To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit
printable ASCII character sequences, e.g. <p\textbackslash> for [ɸ]. Two
problems with SAMPA are that (i) it is only a partial encoding of the IPA and
(ii) it encodes different languages in separate data tables, instead of a
universal alphabet, like IPA. SAMPA tables are derived from phonemes appearing
in several European languages that were developed as part of a European
Commission-funded project to address technical problems like electronic mail
exchange (what is now simply called email). SAMPA is essentially a hack to work
around displaying IPA characters, but it provided speech technology and other
fields a basis that has been widely adopted and used in code. So, SAMPA was a
collection of tables to be compared, instead of a large universal table
representing all languages. An extended version of SAMPA, called X-SAMPA, set
out to include every symbol in the IPA chart including all diacritics
\citep{WellsND}. X-SAMPA was considered more universally applicable because it
consisted of one table that encoded the set of characters that represented
phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used
for speech technology and as an encoding system in computational linguistics.
Eventually, ASCII-encoding of the IPA became deprecated through the advent of
the Unicode Standard. Note however that many popular software packages used for
linguistic analyses still require ASCII input, e.g.~RuG/L04 and
SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and
\url{http://www.splitstree.org/}, respectively}

There are a few pitfalls to be aware of when using the Unicode Standard
to encode IPA. As we have said before, from a linguistic perspective it might
look like the Unicode Consortium is making incomprehensible decisions, but it is
important to realize that the consortium has tried and is continuing to try to
be as consistent as possible across a wide range of use cases, and it does place
linguistic traditions above other orthographic possibilities. In general, we
strongly suggest linguists not to complain about any decisions in the Unicode
Standard, but to try and understand the rationale of the Unicode Consortium
(which in our experience is almost always well-conceived) and devise ways to
work with any unexpected behavior. Many of the current problems derive from the
fact that the IPA is clearly historically based on the Latin script, but
different enough from most other Latin-based writing systems to warrant special
attention. This ambivalent status of the IPA glyphs (partly Latin, partly
special) is unfortunately also attested in the treatment of IPA in the Unicode
Standard. In retrospect, it might have been better to consider the IPA (and
other transcription systems) to be a special or new kind of script within the
Unicode Standard, and treat the obvious similarity to Latin glyphs as a
historical relic. All IPA glyphs would then have their own code points, instead
of the current situation in which some IPA glyphs have special code points,
while others are treated as being identical to the regular Latin characters.
Yet, the current situation, however unfortunate, is unlikely to change, so as
linguists we will have to learn to deal with the specific pitfalls of IPA within
the Unicode Standard. In this section, we will describe these pitfalls in some
detail.
